String Constraints with Regex-Counting and String-Length Solved Efficiently

### 2023-08-02 2023-1-19/sections/preliminaries.tex
grammerly check

We write $\Nat$ and $\Int$ for the sets of natural and integer numbers, respectively. For $n \in \Nat$ with $n \ge 1$, $[n]$ denotes $\{1, \ldots, n\}$; for $m,n \in \Nat$ with $m \le n$,  $[m, n]$ denotes $\{ i \in \Nat \mid m \le i \le n \}$. Throughout the paper, $\Sigma$ is a finite alphabet, ranged over by $a,b,\ldots$.

@@

%\tl{to put in a different place later} Note that our decision procedure is applicable to general transducers as well, however, the EXPSPACE complexity bound is not, because the distributive property $f^{-1}(L_1\cap L_2)= f^{-1}(L_1)\cap f^{-1}(L_2)$ for regular languages $L_1, L_2$ only holds for functional transducers $f$.

@@

%In this paper, we consider logics involving two data-types, i.e., the string data-type and the integer data-type. We will use $u, v, \dots$ to denote string constants,  $c, d,\dots$ to denote integer constants, $x, y, \dots$ to denote string variables, and $i, j, \dots$ to denote  integer variables.

\medskip
\noindent 
\emph{Linear integer arithmetic and Parikh images.}
We will also use standard quantifier-free/existential \emph{linear integer arithmetic} (LIA) formulas, which are typically ranged over by $\phi, \varphi$, etc. 
For a set $\mathfrak{X}$ of variables, we use $\Phi(\mathfrak{X})$ to denote the set of quantifier-free LIA formulas whose free variables are from $\mathfrak{X}$. 
For convenience, we use $\varphi(\vec{\anivar})$ with $\myvec{\anivar} = (\anivar_1, \cdots, \anivar_k)$ to denote an LIA formula $\varphi$ whose free variables are from $\{\anivar_1, \cdots, \anivar_k\}$. For an LIA formula $\varphi(\vec{\anivar})$ with $\myvec{\anivar} = (\anivar_1, \cdots, \anivar_k)$ and $\myvec{t} = (t_1, \cdots, t_k)$ be a tuple of integer terms, we use $\varphi[\myvec{t}/\myvec{\anivar}]$ to denote the formula obtained by replacing (simultaneously) $\anivar_i$ with $t_i$ for every $i \in [k]$.
%For convenience, we usually assume that a set $X$ of variables is a vector. 
%, which is essentially Presburger

>>>>

We write $\Nat$ and $\Int$ for the sets of natural and integer numbers. For $n \in \Nat$ with $n \ge 1$, $[n]$ denotes $\{1, \ldots, n\}$; for $m,n \in \Nat$ with $m \le n$,  $[m, n]$ denotes $\{ i \in \Nat \mid m \le i \le n \}$. Throughout the paper, $\Sigma$ is a finite alphabet, ranged over by $a,b,\ldots$.

@@

%\tl{to put in a different place later} Note that our decision procedure also applies to general transducers. However, the EXPSPACE complexity bound is not because the distributive property $f^{-1}(L_1\cap L_2)= f^{-1}(L_1)\cap f^{-1}(L_2)$ for regular languages $L_1, L_2$ only holds for functional transducers $f$.

@@

%In this paper, we consider logic involving two data types, i.e., the string and integer data types. We will use $u, v, \dots$ to denote string constants,  $c, d,\dots$ to denote integer constants, $x, y, \dots$ to denote string variables, and $i, j, \dots$ to denote integer variables.

\medskip
\noindent 
\emph{Linear integer arithmetic and Parikh images.}
We will also use standard quantifier-free/existential \emph{linear integer arithmetic} (LIA) formulas, which typically range over by $\phi, \varphi$. 
For a set $\mathfrak{X}$ of variables, we use $\Phi(\mathfrak{X})$ to denote the set of quantifier-free LIA formulas whose free variables are from $\mathfrak{X}$. 
For convenience, we use $\varphi(\vec{\anivar})$ with $\myvec{\anivar} = (\anivar_1, \cdots, \anivar_k)$ to denote an LIA formula $\varphi$ whose free variables are from $\{\anivar_1, \cdots, \anivar_k\}$. For an LIA formula $\varphi(\vec{\anivar})$ with $\myvec{\anivar} = (\anivar_1, \cdots, \anivar_k)$ and $\myvec{t} = (t_1, \cdots, t_k)$ be a tuple of integer terms, we use $\varphi[\myvec{t}/\myvec{\anivar}]$ to denote the formula obtained by replacing (simultaneously) $\anivar_i$ with $t_i$ for every $i \in [k]$.
%For convenience, we assume a set $X$ of variables is a vector. 
%, which is essentially Presburger

### 2023-08-02 2023-1-19/sections/recl.tex
grammerly check

%\subsubsection{Syntax}
In the sequel, we define string constraints with regex-counting and string-length function (abbreviated as RECL). The syntax of RECL is defined by the rules in Figure~\ref{fig:syntax}, where $x$ is a string variable, $\anivar$ is an integer variable, $a$ is a character from an alphabet $\Sigma$, and $m, n$ are integer constants. 
A RECL formula $\varphi$ is a conjunction of atomic formulas of the form $x \in \regex$ or $t_1\ \op\ t_2$, where $\regex$ is a regular expression,  
$t_1$ and $t_2$ are integer terms, and $\op \in \{=, \neq, \le, \ge, <, >\}$. Atomic formulas of the form $x \in \regex$ are called \emph{regular membership} constraints and atomic formulas of the form $t_1\ \op\ t_2$ are called \emph{length} constraints. 
%
A regular expression $\regex$ is built from $\emptyset$, the empty string $\epsilon$, and characters $a$ by using concatenation  $\cdot$, union $+$, Kleene star $^*$, intersection $\cap$, complement $\bar{\mbox{ }}$, difference $\setminus$, counting $^{\{m,n\}}$ or $^{\{m,\infty\}}$. An integer term is built from constants $n$, variables $\anivar$, and string lengths $|x|$, by operators $+$ and $-$.
% 
%is a quantifier-free formula that can be a regular membership $x\in\regex$, a quantifier-free Presburger formula $\alpha_1 \leq \alpha_2$, a negation of a formula, or a disjunction of two formulas. $\alpha$ is an integer term which can be an integer constant $m$, integer variable $i$, the length of the string term $|s|$, the minus of an integer term, and the plus of two integer terms. The regular expression $\regex$ is built on empty string $\epsilon$, constant letter $a\in \Sigma$. The supported regex operations involve concatenation $\cdot$, disjunction $+$, intersection $\times$, closure $*$, complement $C$, and bounded repetition $\{m, n\}$. We use $term(\varphi)$ to denote the set of terms and $strvar(\varphi)$ to denote the string variables occurring in $\varphi$. The \emph{literals} of $\varphi$ are $x\in \regex$ and $\alpha_1 \leq \alpha_2$. $x\in \regex$ is called \emph{regular literals} and $\alpha_1 \leq \alpha_2$ is called \emph{linear literals}.
\begin{figure}[h]
\vspace{-3mm}
  \centering
%  \begin{tabular}{l r}
$ \begin{array}{l l l r}
    \varphi & ::= & x\in \regex \mid t_1\  \op\ t_2 \mid  \varphi \wedge \varphi                                              & \mbox{formulas}            \\
    \regex &::= & \emptyset \mid \epsilon \mid a \mid \regex\cdot \regex \mid \regex+\regex \mid \regex^* \mid \regex \cap \regex \mid \overline{\regex} \mid \regex \setminus \regex \mid \regex^{\{m,n\}} \mid \regex^{\{m,\infty\}} \mid (e) & \mbox{regexes} \\
    t &::= & n \mid \anivar \mid  |x| \mid t - t \mid t + t                                                                    & \mbox{integer terms}
    \end{array}
  $
%  \end{tabular}
  \caption{Syntax of RECL }\label{fig:syntax}
\vspace{-3mm}
\end{figure}
Moreover, for $S \subseteq \Sigma$ with $S = \{a_1, \cdots, a_k\}$, we use $S$ as an abbreviation of $a_1 + \cdots + a_k$.

%The definition of the semantics of RECL is standard. Nevertheless, for readibility, we include the definition of its semantics here.

@@

Let $\varphi$ be a RECL formula and $\svars(\varphi)$ (resp. $\ivars(\varphi)$) denote the set of string (resp. integer) variables occurring in $\varphi$. 
%
Then the semantics of  $\varphi$ is defined with respect to a mapping $\theta: \svars(\varphi) \rightarrow \Sigma^* \uplus \ivars(\varphi) \rightarrow \Int$ (where $\uplus$ denotes the disjoint union). 
Note that the mapping $\theta$ can be naturally extended to the set of integer terms. For instance, $\theta(|x|) =|\theta(x)|$, $\theta(t_1 + t_2) = \theta(t_1) + \theta(t_2)$. 
A  mapping $\theta$ is said to satisfy $\varphi$, denoted by $\theta \models \varphi$, if one of the following holds: 
%\begin{itemize}
%\item 
$\varphi \equiv x \in \regex$ and $\theta(x) \in \Lang(\regex)$, 
%
%\item $\varphi \equiv x \not \in \regex$ and $\theta(x) \not \in \Lang(\regex)$, 
%
%\item 
$\varphi \equiv t_1\ \op\ t_2$ and $\theta(t_1)\ \op\ \theta(t_2)$, 
%
%\item 
$\varphi \equiv \varphi_1 \wedge \varphi_2$ and $\theta \models \varphi_1$ and $\theta \models \varphi_2$.
%\end{itemize}
A RECL formula $\varphi$ is said to be satisfiable if there is a mapping $\theta$ such that $\theta \models \varphi$. The satisfiability problem for RECL (abbreviated as $\reclsat$) is to decide whether a given RECL formula $\varphi$ is satisfiable.

>>>>

%\subsubsection{Syntax}
In the sequel, we define string constraints with regex-counting and string-length functions (abbreviated as RECL). The syntax of RECL is defined by the rules in Figure~\ref{fig:syntax}, where $x$ is a string variable, $\anivar$ is an integer variable, $a$ is a character from an alphabet $\Sigma$, and $m, n$ are integer constants. 
A RECL formula $\varphi$ is a conjunction of atomic formulas of the form $x \in \regex$ or $t_1\ \op\ t_2$, where $\regex$ is a regular expression,  
$t_1$ and $t_2$ are integer terms, and $\op \in \{=, \neq, \le, \ge, <, >\}$. Atomic formulas of the form $x \in \regex$ are called \emph{regular membership} constraints, and atomic formulas of the form $t_1\ \op\ t_2$ are called \emph{length} constraints. 
%
A regular expression $\regex$ is built from $\emptyset$, the empty string $\epsilon$, and characters $a$ by using concatenation  $\cdot$, union $+$, Kleene star $^*$, intersection $\cap$, complement $\bar{\mbox{ }}$, difference $\setminus$, counting $^{\{m,n\}}$ or $^{\{m,\infty\}}$. An integer term is built from constants $n$, variables $\anivar$, and string lengths $|x|$, by operators $+$ and $-$.
% 
%is a quantifier-free formula that can be a regular membership $x\in\regex$, a quantifier-free Presburger formula $\alpha_1 \leq \alpha_2$, a negation of a formula, or a disjunction of two formulas. $\alpha$ is an integer term which can be an integer constant $m$, integer variable $i$, the length of the string term $|s|$, the minus of an integer term, and the plus of two integer terms. The regular expression $\regex$ is built on empty string $\epsilon$, constant letter $a\in \Sigma$. The supported regex operations involve concatenation $\cdot$, disjunction $+$, intersection $\times$, closure $*$, complement $C$, and bounded repetition $\{m, n\}$. We use $term(\varphi)$ to denote the set of terms and $strvar(\varphi)$ to denote the string variables occurring in $\varphi$. The \emph{literals} of $\varphi$ are $x\in \regex$ and $\alpha_1 \leq \alpha_2$. $x\in \regex$ is called \emph{regular literals} and $\alpha_1 \leq \alpha_2$ is called \emph{linear literals}.
\begin{figure}[h]
\vspace{-3mm}
  \centering
%  \begin{tabular}{l r}
$ \begin{array}{l l l r}
    \varphi & ::= & x\in \regex \mid t_1\  \op\ t_2 \mid  \varphi \wedge \varphi                                              & \mbox{formulas}            \\
    \regex &::= & \emptyset \mid \epsilon \mid a \mid \regex\cdot \regex \mid \regex+\regex \mid \regex^* \mid \regex \cap \regex \mid \overline{\regex} \mid \regex \setminus \regex \mid \regex^{\{m,n\}} \mid \regex^{\{m,\infty\}} \mid (e) & \mbox{regexes} \\
    t &::= & n \mid \anivar \mid  |x| \mid t - t \mid t + t                                                                    & \mbox{integer terms}
    \end{array}
  $
%  \end{tabular}
  \caption{Syntax of RECL }\label{fig:syntax}
\vspace{-3mm}
\end{figure}
Moreover, for $S \subseteq \Sigma$ with $S = \{a_1, \cdots, a_k\}$, we use $S$ as an abbreviation of $a_1 + \cdots + a_k$.

%The definition of the semantics of RECL is standard. Nevertheless, for readability, we include the definition of its semantics here.

@@

Let $\varphi$ be a RECL formula and $\svars(\varphi)$ (resp. $\ivars(\varphi)$) denote the set of string (resp. integer) variables occurring in $\varphi$. 
%
Then the semantics of  $\varphi$ is defined with respect to a mapping $\theta: \svars(\varphi) \rightarrow \Sigma^* \uplus \ivars(\varphi) \rightarrow \Int$ (where $\uplus$ denotes the disjoint union). 
Note that the mapping $\theta$ can naturally extend to the set of integer terms. For instance, $\theta(|x|) =|\theta(x)|$, $\theta(t_1 + t_2) = \theta(t_1) + \theta(t_2)$. 
A  mapping $\theta$ is said to satisfy $\varphi$, denoted by $\theta \models \varphi$, if one of the following holds: 
%\begin{itemize}
%\item 
$\varphi \equiv x \in \regex$ and $\theta(x) \in \Lang(\regex)$, 
%
%\item $\varphi \equiv x \not \in \regex$ and $\theta(x) \not \in \Lang(\regex)$, 
%
%\item 
$\varphi \equiv t_1\ \op\ t_2$ and $\theta(t_1)\ \op\ \theta(t_2)$, 
%
%\item 
$\varphi \equiv \varphi_1 \wedge \varphi_2$ and $\theta \models \varphi_1$ and $\theta \models \varphi_2$.
%\end{itemize}
A RECL formula $\varphi$ is satisfiable if there is a mapping $\theta$ such as $\theta \models \varphi$. The satisfiability problem for RECL (which is abbreviated as $\reclsat$) is deciding whether a given RECL formula $\varphi$ is satisfiable.
